---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyverse)
```


* Coerce (force) a data frame to a tibble with `as_tibble()`
```{r}
as_tibble(iris)
```

You can create a new tibble from individual vectors with `tibble()`. `tibble()` will automatically recycle inputs of length 1, and allows you to refer to variables that you just created.
```{r}
tibble(
  x = 1:5, 
  y = 1, 
  z = x ^ 2 + y
)
```

## Tibbles vs data.frame

* There are two main differences in the usage of a tibble vs. a classic data.frame: printing and subsetting.

### Printing

Tibbles have a refined print method that shows only the first 10 rows, and all the columns that fit on screen. 

But sometimes you need more output than the default display. There are a few options that can help.

1. Explicitly `print()` the data frame and control the number of rows (n) and the width of the display. `width = Inf` will display all columns:

```{r}
nycflights13::flights %>% 
print(n = 20, width = 10)
```
### Subsetting

If you want to pull out a single variable, you need some new tools, $ and [[. [[ can extract by name or position; $ only extracts by name but is a little less typing.

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

df$x
```
To use these in a pipe, you’ll need to use the special placeholder `.`:
```{r}
df %>% .$x
```

## Interacting with older code

Some older functions don’t work with tibbles. If you encounter one of these functions, use `as.data.frame()` to turn a tibble back to a data.frame:

```{r}
class(as.data.frame(df))
```

## Exercise

1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).

```{r}
# class(mtcars)
is_tibble(mtcars)
```

2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviors cause you frustration?

Answer: 
a. In data frame, the $ operator will match any column name that starts with the name following it.
b. With data.frames, with [ the type of object that is returned differs on the number of columns. If it is one column, it won’t return a data.frame, but instead will return a vector.

```{r}
df <- data.frame(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```
```{r}
df <- tibble(abc = 1, xyz = "a")
df$x
df[, "xyz"]
df[, c("abc", "xyz")]
```

3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?

Answer: use df[[var]] but cannot use df$var.



4. Practice referring to non-syntactic names in the following data frame by:

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```

* Extracting the variable called 1.
* Plotting a scatterplot of 1 vs 2.
* Creating a new column called 3 which is 2 divided by 1.
* Renaming the columns to one, two and three.


```{r}
annoying$`1`
```
```{r}
ggplot(data=annoying, mapping=aes(x=`1`, y=`2`)) + geom_point()
```
```{r}
annoying <- mutate(annoying, `3` = `2` / `1`)
```

```{r}
annoying <- rename(annoying, one=`1`, two=`1`, three=`3`)
glimpse(annoying)
```

5. What does `tibble::enframe()` do? When might you use it?

Answer: The function tibble::enframe() converts named vectors to a data frame with names and values

```{r}
enframe(c(a = 1, b = 2, c = 3))
```

