---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
```


## 1.3 Parsing a Vector

`parse_*()` functions take a character vector and return a more specialised vector like a logical, integer, or date:

```{r}
str(parse_logical(c("TRUE", "FALSE", "NA")))
#>  logi [1:3] TRUE FALSE NA
str(parse_integer(c("1", "2", "3")))
#>  int [1:3] 1 2 3
str(parse_date(c("2010-01-01", "1979-10-14")))
#>  Date[1:2], format: "2010-01-01" "1979-10-14"
```

### 1.3.1 Numbers 

When parsing numbers, the most important option is the character you use for the decimal mark. You can override the default value of . by creating a new locale and setting the decimal_mark argument:


```{r}
parse_double("1.23")
#> [1] 1.23
parse_double("1,23", locale = locale(decimal_mark = ","))
#> [1] 1.23
```

To ignore non-numeric characters before and after the number, we can use `parse_number()`;

```{r}
parse_number("$100")
#> [1] 100
parse_number("20%")
#> [1] 20
parse_number("It cost $123.45")
#> [1] 123.45
```

```{r}
parse_number("123.456.789")
# > [1] 123.456
parse_number("123.456.789", locale = locale(grouping_mark = "."))
# > [1] 123456789
```
## 1.3.3 Factors

R uses factors to represent categorical variables that have a known set of possible values. Give `parse_factor()` a vector of known levels to generate a warning whenever an unexpected value is present:

```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```

### 1.3.4 Dates, date-times, and times

* `parse_datetime()` expects an ISO8601 date-time; ISO8601 is an international standard in which the components of a date are organised from biggest to smallest: year, month, day, hour, minute, second.

```{r}
parse_datetime("2010-10-01T2010")
#> [1] "2010-10-01 20:10:00 UTC"
# If time is omitted, it will be set to midnight
parse_datetime("20101010")
#> [1] "2010-10-10 UTC"
```

* `parse_date()` expects a four digit year, a `-` or `/`, the month, a `-` or `/`, then the day:

```{r}
parse_date("2010-10-01")
```

* 'parse_time()' expects the hour, :, minutes, optionally : and seconds, and an optional am/pm specifier:

```{r}
library(hms)
parse_time("01:10 am")
#> 01:10:00
parse_time("20:10:01")
#> 20:10:01
```



## 1.4 Parsing a file 

```{r}
challenge <- read_csv(readr_example("challenge.csv"))

```
```{r}
problems(challenge)
```


```{r}
challenge <- read_csv(
  readr_example("challenge.csv"), 
  col_types = cols(
    x = col_double(),
    y = col_date()
  )
)
tail(challenge)
```

## 1.5 Try to read an actual file

```{r}
algae_data <- read_csv('data/log_1_chunk_1.csv')
```

