---
title: "R Notebook"
output: html_notebook
---



```{r}
library(tidyverse)
```


## 1.2 Tidy data

There are three interrelated rules which make a dataset tidy:

1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

```{r}
table1
```



## 1.3 Pivoting

### 1.3.1 Longer

A common problem is a dataset where some of the column names are not names of variables, but values of a variable. Take `table4a`: the column names `1999` and `2000` represent values of the year variable, the values in the `1999` and `2000` columns represent values of the cases variable, and each row represents two observations, not one.

```{r}
table4a
```

To tidy a dataset like this, we need to __pivot__ the offending columns into a new pair of variables with function `pivot_longer()`.

`pivot_longer()` makes datasets longer by increasing the number of rows and decreasing the number of columns.

```{r}
table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

```

We can use pivot_longer() to tidy table4b in a similar fashion. The only difference is the variable stored in the cell values:

__original table 4b:__

```{r}
table4b
```
__New table 4b:__

```{r}
table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = 'year', values_to = 'population')
```

To combine the tidied versions of table4a and table4b into a single tibble, we need to use `dplyr::left_join()`;

```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = 'year', values_to = 'population')

left_join(tidy4a, tidy4b)
```

### 1.3.2 Wider

`pivot_wider()` is the opposite of `pivot_longer()`. You use it when an observation is scattered across multiple rows. 
```{r}
table2
```
```{r}
table2 %>%
  pivot_wider(names_from = type, values_from = count)
```


## 1.4 Seperating and Uniting

Motivation: table3 have one column (`rate`) that contains two variables (`cases` and `population`)


### 1.4.1 Seperate

* `separate()` pulls apart one column into multiple columns, by splitting wherever a separator character appears. 
* By default, `separate()` will split values wherever it sees a non-alphanumeric character
* Look carefully at the column types: you’ll notice that cases and population are character columns. This is the default behaviour in separate(): it leaves the type of the column as is. We can ask separate() to try and convert to better types using `convert = TRUE`:

```{r}
table3 %>% 
  separate(rate, into=c("cases", "population"))
```

```{r}
table3 %>% 
  separate(rate, into=c("cases", "population"), convert=TRUE)
```

You can also pass a vector of integers to sep. separate() will interpret the integers as positions to split at. 
* Positive values start at 1 on the far-left of the strings; 
* negative value start at -1 on the far-right of the strings.


```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```

### 1.4.2 Unite
`unite()` is the inverse of `separate()`: it combines multiple columns into a single column.



## 1.5 Missing Values

Changing the representation of a dataset brings up an important subtlety of missing values. Surprisingly, a value can be missing in one of two possible ways:

* Explicitly, i.e. flagged with NA.
* Implicitly, i.e. simply not present in the data.


```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```

There are two missing values in this dataset:

1. The return for the fourth quarter of 2015 is explicitly missing, because the cell where its value should be instead contains NA.

2. The return for the first quarter of 2016 is implicitly missing, because it simply does not appear in the dataset.

We can make those implicit missing values explicit by changing the way that the data is presented. 

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return)
```
Because these explicit missing values may not be important in other representations of the data, you can set `values_drop_na = TRUE` in `pivot_longer()` to turn explicit missing values implicit:

```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>%
  pivot_longer(cols = c(`2015`, `2016`), names_to = "year", values_to = "return", values_drop_na = TRUE)
```

Another important tool for making missing values explicit in tidy data is `complete()`:

* `complete()` takes a set of columns, and finds all unique combinations. It then ensures the original dataset contains all those values, filling in explicit `NAs` where necessary.

```{r}
stocks %>%
    complete(year, qtr)
```


There’s one other important tool that you should know for working with missing values. Sometimes when a data source has primarily been used for data entry, missing values indicate that the previous value should be carried forward:

```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
```

We can fill in these missing values with `fill()`, with the targeted columns to be replaced by the most recent non-missing value. 


```{r}
treatment %>% 
  fill(person)
```

