---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyverse)
library(nycflights13)
```

## 1.1 nycflights13

There are four tibbles that are related to the `flights` table that you used in data transformation. 

1. `airlines` lets you look up the full carrier name from its abbreviated code:

```{r}
airlines
```

2. `airports` gives information about each airport, identified by the `faa` airport code:
```{r}
airports
```

3. `planes` gives information about each plane, identified by its `tailnum`:

```{r}
planes
```

`weather` gives the weather at each NYC airport for each hour:

```{r}
weather
```

## 1.2 Keys

There are two types of keys:

1. A __primary key__ uniquely identifies an observation in its own table. For example, `planes$tailnum` is a primary key because it uniquely identifies each plane in the planes table.

2. A __foreign key__ uniquely identifies an observation in another table. For example, `flights$tailnum` is a foreign key because it appears in the flights table where it matches each flight to a unique plane.


Once you’ve identified the primary keys in your tables, it’s good practice to verify that they do indeed uniquely identify each observation. One way to do that is to count() the primary keys and look for entries where n is greater than one:

```{r}
planes %>%
  count(tailnum) %>%
  filter(n > 1)

weather %>% 
  count(year, month, day, hour, origin) %>% 
  filter(n > 1)
```

If a table lacks a primary key, it’s sometimes useful to add one with `mutate()` and `row_number()`. That makes it easier to match observations if you’ve done some filtering and want to check back in with the original data. This is called a __surrogate key__.


## 1.3 Mutating Joins

Imagine you want to add the full airline name to the flights2 data. You can combine the `airlines` and `flights2` data frames with `left_join()`:

```{r}
flights2 <- flights %>% 
  select(year:day, hour, origin, dest, tailnum, carrier)
# view(flights2)


flights2 %>% 
  select(-origin, -dest) %>%  # other than origin and dest columns
  left_join(airlines, by = "carrier")
```

## 1.4 Filtering Joins

Filtering joins match observations in the same way as mutating joins, but affect the observations, not the variables. There are two types:

* `semi_join(x, y)` __keeps__ all observations in `x` that have a match in `y`.
* `anti_join(x, y)` __drops__ all observations in `x` that have a match in `y`.
  *  For example, when connecting flights and planes, you might be interested to know that there are many flights that don’t have a match in planes.


## 1.5 Set Operations

All these operations work with a complete row, comparing the values of every variable. These expect the x and y inputs to have the same variables, and treat the observations like sets:

* intersect(x, y): return only observations in both x and y.
* union(x, y): return unique observations in x and y.
* setdiff(x, y): return observations in x, but not in y.



