---
title: "R Notebook"
output: html_notebook
---



```{r}
library(tidyverse)
```

## An example data set - General Social Survey

```{r}
gss_cat
```

When factors are stored in a tibble, you can’t see their levels so easily. One way to see them is with `count()`:

```{r}
gss_cat %>%
  count(race)
```
By default, ggplot2 will drop levels that don’t have any values. You can force them to display with:

```{r}
ggplot(gss_cat, aes(race)) + geom_bar() + scale_x_discrete(drop = FALSE)
```

## 1.2 Modifying factor order

It’s often useful to change the order of the factor levels in a visualisation. For example, imagine you want to explore the average number of hours spent watching TV per day across religions:

```{r}
relig_summary <- gss_cat %>%
      group_by(relig) %>%
      summarise(
        age = mean(age, na.rm = TRUE),
        tvhours = mean(tvhours, na.rm=TRUE),
        n = n()
      )

ggplot(relig_summary, aes(x=tvhours, y=relig)) + geom_point()
```
It is difficult to interpret this plot because there’s no overall pattern. We can improve it by reordering the levels of relig using `fct_reorder()`. `fct_reorder()` takes three arguments:

1. `f`, the factor whose levels you want to modify.
2. `x`, a numeric vector that you want to use to reorder the levels.
3. Optionally, `fun`, a function that’s used if there are multiple values of `x` for each value of `f`. The default value is median.

```{r}
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) + geom_point()
```

A better way to write:

```{r}
relig_summary %>%
  mutate(relig = fct_reorder(relig, tvhours)) %>%
  ggplot(aes(tvhours, relig)) + 
  geom_point()
```

Another type of reordering is useful when you are colouring the lines on a plot. `fct_reorder2()` reorders the factor by the y values associated with the largest x values. This makes the plot easier to read because the line colours line up with the legend.

```{r}
by_age <- gss_cat %>%
  filter(!is.na(age)) %>%
  count(age, marital) %>%
  group_by(age) %>%
  mutate(prop = n / sum(n))


ggplot(by_age, aes(age, prop, color = marital)) + geom_line(na.rm = TRUE)

ggplot(by_age, aes(age, prop, color = fct_reorder2(marital, age, prop))) +  geom_line() +
  labs(colour = "marital")
```

## 1.3 Modifying factor levels

More powerful than changing the orders of the levels is changing their values. This allows you to clarify labels for publication, and collapse levels for high-level displays. The most general and powerful tool is `fct_recode()`.

Original partyid:

```{r}
gss_cat %>% count(partyid)
```

After tweaking by `fct_recode` with a parallel construction:
```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
  )) %>%
  count(partyid)
```


To combine groups, you can assign multiple old levels to the same new level:
```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat",
    "Other"                 = "No answer",
    "Other"                 = "Don't know",
    "Other"                 = "Other party"
  )) %>%
  count(partyid)
```
If you want to collapse a lot of levels, `fct_collapse()` is a useful variant of `fct_recode()`. For each new variable, you can provide a vector of old levels:
(combining multiple levels together)

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

