---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyverse)

library(lubridate)
library(nycflights13)
```
## 1.1 Creating date/times

There are three types of date/time data that refer to an instant in time:

1. A date. Tibbles print this as `<date>`. (will be covered)

2. A time within a day. Tibbles print this as `<time>`. (will not be covered)

3. A date-time is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as `<dttm>`. (will be covered)


There are serveal ways to create a date/time: 

* To get the current date or date-time you can use today() or now():

```{r}

today()
now()

```
* From a string.
* From individual date-time components.
* From an existing date/time object.


### 1.1.1 From Strings

We can use the helpers provided by __lubridate__. They automatically work out the format once you specify the order of the component. To use them, identify the order in which year, month, and day appear in your dates;

```{r}
ymd("2017-01-31")
#> [1] "2017-01-31"
mdy("January 31st, 2017")
#> [1] "2017-01-31"
dmy("31-Jan-2017")
#> [1] "2017-01-31"
```
These functions also take unquoted numbers. 
```{r}
ymd(20170131)
```

To create a date-time, we than add an underscore and one or more of “h”, “m”, and “s” to the name of the parsing function:

```{r}
ymd_hms("2017-01-31 20:11:59")
#> [1] "2017-01-31 20:11:59 UTC"
mdy_hm("01/31/2017 08:01")
#> [1] "2017-01-31 08:01:00 UTC"
```

### 1.1.2 From individual components
Instead of a single string, sometimes you’ll have the individual components of the date-time spread across multiple columns. This is what we have in the flights data:

```{r}
flights %>% 
  select(year, month, day, hour, minute)
```
To create a date/time from this sort of input, use `make_date()` for dates, or `make_datetime()` for date-times:

```{r}
flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(departure = make_datetime(year, month, day, hour, minute))


```

