---
title: "R Notebook"
output: html_notebook
---

# Chapter 5 Multivariate Graphs

Multivariate graphs display the relationships among three or more variables. There are two common methods for accommodating multiple variables: grouping and faceting.

```{r}
library(tidyverse)
library(ggplot2)
```

## 5.1 Grouping

In grouping, the values of the first two variables are mapped to the x and y axes. Then additional variables are mapped to other visual characteristics such as color, shape, size, line type, and transparency. Grouping allows you to plot the data for multiple groups in a single graph.

Using the Salaries dataset, let’s display the relationship between yrs.since.phd and salary.
```{r}
data(Salaries, package="carData")

# plot experience vs. salary
ggplot(Salaries, 
       aes(x = yrs.since.phd, 
           y = salary)) +
  geom_point() + 
  labs(title = "Academic salary by years since degree")
```
Next, let’s include the rank of the professor, using color.

```{r}
ggplot(Salaries, 
       aes(x = yrs.since.phd, 
           y = salary,
           color=rank)) +
  geom_point() + 
  labs(title = "cademic salary by rank and years since degree")
```
Finally, let’s add the gender of professor, using the shape of the points to indicate sex. We’ll increase the point size and add transparency to make the individual points clearer.
```{r}
ggplot(Salaries, 
       aes(x = yrs.since.phd, 
           y = salary,
           color=rank,
           shape=sex)) +
  geom_point(size = 3,
             alpha = .6) + 
  labs(title = "Academic salary by rank, sex, and years since degree")
```
Here is a cleaner example. We’ll graph the relationship between years since Ph.D. and salary using the __size of the points__ to indicate years of service. This is called a bubble plot.
```{r}
# plot experience vs. salary 
# (color represents rank and size represents service)
ggplot(Salaries, 
       aes(x = yrs.since.phd, 
           y = salary, 
           color = rank, 
           size = yrs.service)) +
  geom_point(alpha = .6) + 
  scale_y_continuous(label = scales::dollar) +
  labs(title = "Academic salary by rank, years of service, and years since degree")
```

As a final example, let’s look at the yrs.since.phd vs salary and add sex using color and quadratic best fit lines.
```{r}
# plot experience vs. salary with 
# fit lines (color represents sex)
ggplot(Salaries, 
       aes(x = yrs.since.phd, 
           y = salary, 
           color = sex)) +
  geom_point(alpha = .4, 
             size = 3) +
  geom_smooth(se=FALSE, 
              method = "lm", 
              formula = y~poly(x,2), 
              size = 1.5) +
  labs(x = "Years Since Ph.D.",
       title = "Academic Salary by Sex and Years Experience",
       subtitle = "9-month salary for 2008-2009",
       y = "",
       color = "Sex") +
  scale_y_continuous(label = scales::dollar) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal()
```
## 5.2 Faceting

In faceting, a graph consists of several separate plots or small multiples, one for each level of a third variable, or combination of variables. It is easiest to understand this with an example.

```{r}
# plot salary histograms by rank
ggplot(Salaries, aes(x = salary)) +
  geom_histogram(fill = "cornflowerblue",
                 color = "white") +
  facet_wrap(~rank, ncol = 1) +
  labs(title = "Salary histograms by rank")
```
The `facet_wrap` function creates a separate graph for each level of `rank`. The `ncol` option controls the number of columns.


In the next example, two variables are used to define the facets. The format of the `facet_grid` function is

`facet_grid( row variable(s) ~ column variable(s))`

```{r}
# plot salary histograms by rank and sex
ggplot(Salaries, aes(x = salary / 1000)) +
  geom_histogram(color = "white",
                 fill = "cornflowerblue") +
  facet_grid(sex ~ rank) +
  labs(title = "Salary histograms by sex and rank",
       x = "Salary ($1000)")
```
We can also combine grouping and faceting. Let’s use Mean/SE plots and faceting to compare the salaries of male and female professors, within rank and discipline. We’ll use color to distinguish sex and faceting to create plots for rank by discipline combinations.

```{r}
plotdata <- Salaries %>%
  group_by(sex, rank, discipline) %>%
  summarize(n = n(),
            mean = mean(salary),
            sd = sd(salary),
            se = sd / sqrt(n))

# create better labels for discipline
plotdata$discipline <- factor(plotdata$discipline,
                              labels = c("Theoretical",
                                         "Applied"))
# create plot
ggplot(plotdata, 
       aes(x = sex, 
           y = mean,
           color = sex)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean - se, 
                    ymax = mean + se),
                width = .1) +
  scale_y_continuous(breaks = seq(70000, 140000, 10000),
                     label = scales::dollar) +
  facet_grid(. ~ rank + discipline) +
  theme_bw() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(x="", 
       y="", 
       title="Nine month academic salaries by gender, discipline, and rank",
       subtitle = "(Means and standard errors)") +
  scale_color_brewer(palette="Set1")
```

Note:
1. The statement `facet_grid(. ~ rank + discipline)` specifies no row variable (.) and columns defined by the combination of rank and discipline.
2. The `theme_` functions create create a black and white theme and eliminates vertical grid lines and minor horizontal grid lines. 
3. The `scale_color_brewer` function changes the color scheme for the points and error bars.


As a final example, we’ll shift to a new dataset and plot the change in life expectancy over time for countries in the “Americas”. The data comes from the gapminder dataset in the gapminder package. Each country appears in its own facet. The theme functions are used to simplify the background color, rotate the x-axis text, and make the font size smaller.

```{r}
data(gapminder, package = "gapminder")

# Select the Americas data
plotdata <- dplyr::filter(gapminder, 
                          continent == "Americas")

# plot life expectancy by year, for each country
ggplot(plotdata, aes(x=year, y = lifeExp)) +
  geom_line(color="grey") +
  geom_point(color="blue") +
  facet_wrap(~country) + 
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1)) +
  labs(title = "Changes in Life Expectancy",
       x = "Year",
       y = "Life Expectancy") 
```

