---
title: "R Notebook"
output: html_notebook
---

```{r}
library(nycflights13)
library(tidyverse)

arrange(flights, year, month, day)
```

Use desc() to re-order by a column in descending order:


```{r}
arrange(flights, desc(dep_delay))
```

Missing values are always sorted at the end:

```{r}
df <- tibble(x = c(5, 2, NA))
arrange(df, x)
```
__Exercise__:

1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

```{r}
arrange(flights, desc(is.na(dep_time)), dep_time)
```

2. Sort flights to find the most delayed flights. Find the flights that left earliest.
```{r}
# the most delayed left
arrange(flights, desc(arr_delay))
```
```{r}
# the earliest left
arrange(flights, dep_delay)
```
3. Sort flights to find the fastest (highest speed) flights.
```{r}
head(arrange(flights, desc(distance / air_time)))
```

4. Which flights travelled the farthest? Which travelled the shortest?
```{r}
# the farthest
head(arrange(flights, desc(distance)))
```

```{r}
# the shortest
head(arrange(flights, distance))
```