---
title: "R Notebook"
output: html_notebook
---

```{r}
library(nycflights13)
library(tidyverse)
```

It’s not uncommon to get datasets with hundreds or even thousands of variables. In this case, the first challenge is often narrowing in on the variables you’re actually interested in. 

`select()` allows you to rapidly zoom in on a useful subset using operations based on the names of the variables.

```{r}
select(flights, year, month, day)
```

There are a number of helper functions you can use within `select()`:

* `starts_with("abc")`: matches names that begin with “abc”.

* `ends_with("xyz")`: matches names that end with “xyz”.

* `contains("ijk")`: matches names that contain “ijk”.

* `matches("(.)\\1")`: selects variables that match a regular expression. This one matches any variables that contain repeated characters. You’ll learn more about regular expressions in strings.

* `num_range("x", 1:3)`: matches x1, x2 and x3.


To rename a column (variable), we can use `rename()` function.

```{r}
rename(flights, tail_num = tailnum)
```

Another option is to use `select()` in conjunction with the `everything()` helper. This is useful if you have a handful of variables you’d like to move to the start of the data frame. 

Move the interested columns to front. 

```{r}
select(flights, time_hour, air_time, everything())
```

__Exercies__:

1. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from flights.


```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
```

2. What happens if you include the name of a variable multiple times in a `select()` call?

The `select()` call ignores the duplication. Any duplicated variables are only included once, in the first location they appear. The `select()` function does not raise an error or warning or print any message if there are duplicated variables.


3. What does the one_of() function do? Why might it be helpful in conjunction with this vector?

In the most recent versions of dplyr, one_of has been deprecated in favor of two functions: `all_of()` and `any_of()`.

These functions differ in their strictness. The function `all_of()` will raise an error if one of the variable names is not present, while `any_of()` will ignore it.


```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

select(flights, any_of(vars))
```
4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?


The default behavior for contains() is to ignore case, to make it case sensitive, we can add an argument 'ignore.case = FALSE.'

```{r}
select(flights, contains("TIME"))
```
```{r}
select(flights, contains("TIME", ignore.case = FALSE))
```



