---
title: "R Notebook"
output: html_notebook
---

```{r}
library(nycflights13)
library(tidyverse)
```

`mutate()` always adds new columns at the end of your dataset.

```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)

mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)

```
If you only want to keep the new variables, use `transmute()`.

```{r}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```
__Useful creation functions__

* Arithmetic operators: +, -, *, /, ^

* Modular arithmetic: `%/%` (integer division) and `%%` (remainder), where x == y * (x %/% y) + (x %% y)

* Logs: log(), log2(), log10()

* Offsets: lead() and lag() allow you to refer to leading or lagging values. 

```{r}
x <- 1:10
lag(x)
lead(x)
```
* Cumulative and rolling aggregates: R provides functions for running sums, products, mins and maxes: `cumsum()`, `cumprod()`, `cummin()`, `cummax()`; and dplyr provides `cummean()` for cumulative means.

```{r}
cumsum(x)
cummean(x)
```
* Logical comparisons, <, <=, >, >=, !=, and ==;

* Ranking: there are a number of ranking functions, but you should start with `min_rank()`.


__Exercise__

1. Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.
```{r}
time2mins <- function(x) {
  (x %/% 100 * 60 + x %% 100) %% 1440
}
```

```{r}
flights_times <- mutate(flights, 
                        dep_time_mins = time2mins(dep_time),
                        sched_dep_time_mins = time2mins(sched_dep_time)
)

select(
  flights_times, dep_time, dep_time_mins, sched_dep_time,
  sched_dep_time_mins
)

```
2. Compare `air_time` with `arr_time` - `dep_time`. What do you expect to see? What do you see? What do you need to do to fix it?

```{r}
flights_airtime <- mutate(flights, dep_time = time2mins(dep_time), arr_time = time2mins(arr_time), 
                         air_time_diff = air_time - arr_time + dep_time)

nrow(filter(flights_airtime, air_time_diff != 0))
```

3. Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?

```{r}
flights_deptime <-
  mutate(flights,
    dep_time_min = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    sched_dep_time_min = (sched_dep_time %/% 100 * 60 +
      sched_dep_time %% 100) %% 1440,
    dep_delay_diff = dep_delay - dep_time_min + sched_dep_time_min
  )

nrow(filter(flights_deptime, dep_delay_diff != 0))
```

```{r}
ggplot(
  filter(flights_deptime, dep_delay_diff > 0),
  aes(y = sched_dep_time_min, x = dep_delay_diff)
) +
  geom_point()
```
4. Find the 10 most delayed flights using a ranking function. How do you want to handle ties? 


```{r}
rankme <- tibble(
  x = c(10, 5, 1, 5, 5)
)

rankme <- mutate(rankme,
  x_row_number = row_number(x),
  x_min_rank = min_rank(x),
  x_dense_rank = dense_rank(x)
)

arrange(rankme, x)
```

The `min_rank()` function assigns a rank equal to the number of values less than that tied value plus one. In contrast, the `dense_rank()` function assigns a rank equal to the number of distinct values less than that tied value plus one.

```{r}
flights_delay <- mutate(flights, dep_delay_min_rank = min_rank(desc(dep_delay)),
                        dep_delay_row_number = row_number(desc(dep_delay)),
                        dep_delay_dense_rank = dense_rank(desc(dep_delay)))

flights_delay <- filter(flights_delay, !(dep_delay_min_rank > 10 | dep_delay_row_number > 10 | dep_delay_dense_rank > 10))

flights_delay <- arrange(flights_delay, dep_delay_min_rank)
print(flights_delay)
```

There are other ways to solve this problem that do not using ranking functions. To select the top 10, sort values with `arrange()` and select the top values with `slice`:

```{r}
flights_delay_2 <- arrange(flights, desc(dep_delay))
flights_delay_2 <- slice(flights_delay_2, 1:10)
select(flights_delay_2,  month, day, carrier, flight, dep_delay)
```


Alternatively, we could use the `top_n()`.
```{r}
flights_delayed3 <- top_n(flights, 10, dep_delay)
flights_delayed3 <- arrange(flights_delayed3, desc(dep_delay))
select(flights_delayed3, month, day, carrier, flight, dep_delay)
```

5. What does 1:3 + 1:10 return? Why?

```{r}
1:3 + 1:10

```

This is equivalent to the following.
```{r}
c(1 + 1, 2 + 2, 3 + 3, 1 + 4, 2 + 5, 3 + 6, 1 + 7, 2 + 8, 3 + 9, 1 + 10)
```

