---
title: "R Notebook"
output: html_notebook
---

### Function Basics

There are __three__ key steps to creating a new function:

1. You need to pick a __name__ for the function. 

2. You list the inputs, or __arguments__, to the function inside function. Here we have just one argument. If we had more the call would look like function(x, y, z).

3. You place the code you have developed in body of the function, a `{` block that immediately follows function(...).


A rescale function example: 
```{r}

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)  ## range returns the min and max of an array
  (x - rng[1]) / (rng[2] - rng[1])
}

rescale01(c(0, 5, 10))
```
```{r}
range(c(5, 10, 1, 20))
```

### Conditional Execution

An `if` statement allows you to conditionally execute code. It looks like this:

```{r, eval=FALSE}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```

You can chain multiple if statements together: 

```{r, eval=FALSE}
if (this) {
  # do that
} else if (that) {
  # do something else
} else {
  # 
}

```

But if you end up with a very long series of chained if statements, you should consider rewriting. One useful technique is the `switch()` function. It allows you to evaluate selected code based on position or name.

```{r, eval=FALSE}
> function(x, y, op) {
>   switch(op,
>     plus = x + y,
>     minus = x - y,
>     times = x * y,
>     divide = x / y,
>     stop("Unknown op!")
>   )
> }
```

