---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyverse)
```

## 1.1 Iteration

### 1. Write for loops to:
1. Compute the mean of every column in `mtcars`.
```{r}
outputs <- vector("double", ncol(mtcars))
names(outputs) <- names(mtcars)
for (i in names(mtcars)){
  outputs[[i]] <- mean(mtcars[[i]])
}
outputs
```


2. Determine the type of each column in nycflights13::flights.

```{r}
library(nycflights13)
cols_type <- vector("list", ncol(flights))
names(cols_type) <- names(flights)
for (i in names(flights)){
  cols_type[[i]] <- class(flights[[i]])
}
cols_type
```

3. Compute the number of unique values in each column of iris.

```{r}
data("iris")
num_uniques <- vector("double", ncol(iris))
names(num_uniques) <- names(iris)
for (i in seq_along(iris)){
  num_uniques[[i]] <- n_distinct(iris[[i]])
}
num_uniques
```

4. Generate 10 random normals from distributions with means of -10, 0, 10, and 100.


```{r}
n <- 10
mu <- c(-10, 0, 10, 100)

normals <- vector("list", length(mu))
for (i in seq_along(mu)){
  normals[[i]] <- rnorm(n, mean=mu[i])
}
normals

```
### 1.2 Eliminate the for loop in each of the following examples by taking advantage of an existing function that works with vectors:


```{r}
out <- ""
for (x in letters) {
  out <- stringr::str_c(out, x)
}

## Alternative
str_c(letters, collapse="")

```


```{r}
# x <- sample(100)
# sd <- 0
# for (i in seq_along(x)) {
#   sd <- sd + (x[i] - mean(x)) ^ 2
# }
# sd <- sqrt(sd / (length(x) - 1))

## Alternatively
mean_x <- mean(x)
result <- sqrt(sum((x - mean_x)^2) / (length(x) - 1))
result
```

```{r}
x <- runif(100)
out <- vector("numeric", length(x))
out[1] <- x[1]
for (i in 2:length(x)) {
  out[i] <- out[i - 1] + x[i]
}

## equivalent to cumulative sum
all.equal(cumsum(x), out)

```

### 1.3 Combine your function writing and for loop skills: Write a for loop that prints() the lyrics to the children’s song “Alice the camel”.

```{r}
humps <- c("five", "four", "three", "two", "one", "no")
for (i in humps){
  cat(str_c("Alice the camel has ", rep(i, 3), " humps", sep = "", collapse = '\n'), "\n")
  if (i == 'no'){
    cat("Now Alice is a horse.\n")
  }else{
    cat("So go, Alice, go.\n")
  }
  
  cat("\n")
}
```




