---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyverse)
library(tidymodels)
library(rcfss)
library(modelr)
set.seed(123)
```

## Q1: Linear Regression with `scorecard`

### 1.1 What is the relationship between admission rate and cost? Report this relationship using a scatterplot and a linear best-fit line.



```{r}
ggplot(data=scorecard, aes(x=admrate, y=cost)) + geom_point() + geom_smooth(method="lm")
```
 
### 1.2 Estimate a linear regression of the relationship between admission rate and cost, and report your results in a tidy table.

```{r}
linear_model <- lm(cost ~ admrate, data=scorecard, na.action=na.exclude)
```

```{r}
grid <- scorecard %>%
  data_grid(admrate = seq_range(admrate, 20)) %>%
  add_predictions(linear_model, "cost") ## from modelr package

ggplot(scorecard, aes(x = admrate, y = cost)) +  geom_hex(bins = 50) + 
  geom_line(data = grid, colour = 'red', size = 1)
```





```{r}
scorecard_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(cost ~ admrate, data=scorecard)

tidy(scorecard_fit)
```

### 1.3 Estimate a linear regression of the relationship between admission rate and cost, while also accounting for type of college and percent of Pell Grant recipients, and report your results in a tidy table.

```{r}
scorecard_fit <-linear_reg() %>%
  set_engine("lm") %>%
  fit(cost ~ admrate + type + pctpell, data = scorecard)

tidy(scorecard_fit)
```


## Q2 Logistic Regression with `mental_health`

### 2.1 Estimate a logistic regression model of voter turnout with `mhealth` as the predictor. Estimate predicted probabilities and a 95% confidence interval, and plot the logistic regression predictions using ggplot.

```{r}
## convert vote96 to a factor column
mental_health <- rcfss::mental_health %>%
  mutate(vote96 = factor(vote96, labels = c("Not voted", "Voted")))


## estimate moidel
mh_mod <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(vote96 ~ mhealth, data = mental_health)

## generate predicted probabilities + confidence intervals
new_points <- tibble(
  mhealth = seq(
    from = min(mental_health$mhealth),
    to = max(mental_health$mhealth)
  )
)

bind_cols(
  new_points,
  # predicted probabilities
  predict(mh_mod, new_data = new_points, type = "prob"),
  # confidence intervals
  predict(mh_mod, new_data = new_points, type = "conf_int")
) %>%
  # graph the predictions
  ggplot(mapping = aes(x = mhealth, y = .pred_Voted)) +
  geom_pointrange(mapping = aes(ymin = .pred_lower_Voted, ymax = .pred_upper_Voted)) +
  labs(
    title = "Relationship Between Mental Health and Voter Turnout",
    x = "Mental health status",
    y = "Predicted Probability of Voting"
  )

```

### 2.2. Estimate a second logistic regression model of voter turnout using using age and gender (i.e. the `female` column). Extract predicted probabilities and confidence intervals for all possible values of age, and visualize using `ggplot()`.

```{r}
mental_health <- mental_health %>%
  mutate(female, factor(female, labels = c("Male", "Female")))

## estimate model
mh_int_mod <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(vote96 ~ age * female, data = mental_health)

## generate predicted probabilities + confidence intervals
new_points <- expand.grid(
  age = seq(
    from = min(mental_health$age),
    to = max(mental_health$age)
  ),
  female = unique(mental_health$female)
)

bind_cols(
  new_points,
  # predicted probabilities
  predict(mh_int_mod, new_data = new_points, type = "prob"),
  # confidence intervals
  predict(mh_int_mod, new_data = new_points, type = "conf_int")
) %>%
  # graph the predictions
  ggplot(mapping = aes(x = age, y = .pred_Voted, color = female)) +
  # predicted probability
  geom_line(linetype = 2) +
  # confidence interval
  geom_ribbon(mapping = aes(
    ymin = .pred_lower_Voted, ymax = .pred_upper_Voted,
    fill = female
  ), alpha = .2) +
  scale_color_viridis_d(
    end = 0.7, aesthetics = c("color", "fill"),
    name = NULL
  ) +
  labs(
    title = "Relationship Between Age and Voter Turnout",
    x = "Age",
    y = "Predicted Probability of Voting"
  )
```

