---
title: "R Notebook"
output: html_notebook
---



```{r}
library(tidyverse)
```

```{r}
library(rcfss)
data("scorecard")
glimpse(scorecard)
```

### Q1: Generate a data frame of schools with a greater than 40% share of first-generation students

```{r}
first_generation_40 <- scorecard %>%
  filter(firstgen > 0.4, na.rm=TRUE) %>%
  select(unitid, name)
```


### Q2: Generate a data frame with the 10 most expensive colleges in 2018-19 based on net cost of attendance
```{r}
most_expensive_10 <- scorecard %>%
  arrange(desc(netcost)) %>%
  slice(1:10)
```

we can also use `slice_max` in `dplyr` to accomplish the same thing in one line of code.

```{r}
slice_max(.data = scorecard, n = 10, netcost)
```

### Q3 Generate a data frame with the average SAT score for each type of college

```{r}
average_SAT <- scorecard %>%
  group_by(type) %>%
  summarise(mean_SAT = mean(satavg, na.rm =TRUE))
```

### Q4: Calculate for each school how many students it takes to pay the average faculty member’s salary and generate a data frame with the school’s name and the calculated value

let's first check if there are repeated schools or `NA`.
```{r}
n_distinct(scorecard$unitid)
n_distinct(scorecard$name)
any(is.na(scorecard$name))
```

```{r}
count_students <- scorecard %>%
  mutate(num_students = avgfacsal / netcost) %>%
  filter(!is.na(num_students)) %>%
  select(name, num_students) %>%
  mutate_if(is.numeric, scales::number, accuracy=.01) # 1 means only integer, .1, keep one decimal 
  
```


### Q5 Calculate how many private, nonprofit schools have a smaller net cost than the University of Chicago   

Hint: the result should be a data frame with one row for the University of Chicago, and a column containing the requested value.


Report the number as the total number of schools with `row_number()``

```{r}
scorecard %>%
  filter(type  == 'Private, nonprofit') %>%
  arrange(netcost) %>%
  mutate(school_cheaper = row_number() - 1) %>%
  filter(name == 'University of Chicago') %>%
  glimpse
  
```

Report the number as the percentage of schools with `percent_rank`:

```{r}
scorecard %>%
  filter(type == 'Private, nonprofit') %>%
  mutate(netcost_rank = percent_rank(netcost)) %>%
  filter(name=='University of Chicago') %>%
  glimpse()
```

